---
title: "Bayesian Updating Blog - Hecksteden et al (2017)"
author: "Patrick Ward"
date: "11/28/2021"
output: html_document
---

**Reference:** Hecksteden et al. (2017). **A new method to individualize monitoring of muscle recovery in athletes.** Int J Sport Phys Perf; 12: 1137-1142.


## Example Plots for Showing Reference Ranges


```{r}
library(tidyverse)
library(patchwork)
library(zoo)

theme_set(theme_light())

set.seed(4059)
training_session <- 1:40
training_load <- c(rnorm(n = 20, mean = 300, sd = 25), 
                   rnorm(n = 10, mean = 375, sd = 15),
                   rnorm(n = 10, mean = 280, sd = 40))

df <- tibble(training_session, training_load) %>%
  mutate(training_load_scaled = (training_load - mean(training_load)) / sd(training_load),
         roll_avg = rollapply(training_load, width = 10, by = 20, FUN = "mean", fill = NA, align = "right"),
         roll_sd = rollapply(training_load, width = 10, by = 20, FUN = "sd", fill = NA, align = "right")) %>%
  fill(c(roll_avg, roll_sd), .direction = "down")


figA <- ggplot(df, aes(x = training_session, y = training_load_scaled)) +
  geom_rect(aes(ymin = -1,
                ymax =1),
            xmin = 0,
            xmax = Inf,
            fill = "light grey",
            alpha = 0.4) +
  geom_hline(yintercept = 0,
             size = 1.2) +
  geom_line(size = 1.2) +
  geom_point(shape = 21,
             fill = "white",
             size = 4) +
  ggtitle("Figure A: Z-Scores") +
  ylim(-3, 3)


figB <- ggplot(df, aes(x = training_session, y = training_load)) +
  geom_rect(aes(ymin = mean(training_load) - sd(training_load),
                ymax = mean(training_load) + sd(training_load)),
            xmin = 0,
            xmax = Inf,
            fill = "light grey",
            alpha = 0.4) +
  geom_hline(aes(yintercept = mean(training_load)),
             size = 1.2) +
  geom_line(size = 1.2) +
  geom_point(shape = 21,
             fill = "white",
             size = 4) +
  ggtitle("Figure B: Raw Values") +
  ylim(150, 420)

figC <- ggplot(df, aes(x = training_session, y = training_load)) +
  geom_ribbon(aes(ymin = roll_avg - roll_sd,
                ymax = roll_avg + roll_sd),
            fill = "light grey",
            alpha = 0.6) +
  geom_line(aes(y = roll_avg),
             size = 1.2) +
  geom_line(size = 1.2) +
  geom_point(shape = 21,
             fill = "white",
             size = 4) +
  ggtitle("Figure C: Rolling Mean & SD") +
  ylim(150, 420)


(figA | figB) / (figC | plot_spacer())

```


## Bayesian Updating for Reference Ranges

**Create data frame of tests**

```{r}
df <- tibble(
  test = 1:5,
  value = c(90,110,97,95,102)
)

# Take the log of each value
df$ln_value <- log(df$value)
df

```


**Set Priors (in log format)**

```{r}
## priors
prior_mu <- 5.527
prior_sd <- 0.661
prior_repeated_measure_sd <- 0.504
```


**Create functions to update the prior parameters**

```{r}
posterior_mu <- function(prior_mu, prior_sd, prior_repeated_measure_sd, obs_value){
  
  numerator <- prior_repeated_measure_sd^2 * prior_mu + prior_sd^2 * obs_value
  denominator <- prior_repeated_measure_sd^2 + prior_sd^2
  
  post_mu <- numerator / denominator
  return(post_mu)
  
  }

posterior_sd <- function(prior_repeated_measure_sd, prior_sd, test_num){
  
  post_var <- 1 / ((test_num - 1 + 1) * 1/prior_repeated_measure_sd^2 + 1/prior_sd^2) 
  post_sd <- sqrt(post_var)
  return(post_sd)
  
}

```

**Update the mean and standard deviation with the first observation in the data**

```{r}
post_mu <- posterior_mu(prior_mu = prior_mu,
             prior_sd = prior_sd,
             prior_repeated_measure_sd = prior_repeated_measure_sd,
             obs_value = 4.5)

post_sd <- posterior_sd(prior_repeated_measure_sd = prior_repeated_measure_sd,
             prior_sd = prior_sd,
             test_num = 1)

post_mu
post_sd

```

**Calculate the 95% CI**

```{r}
## 95% CI
high95 <- post_mu + 1.96 * post_sd
low95 <- post_mu - 1.96 * post_sd

high95
low95

## Exponentiate back to the normal scale
exp(high95)
exp(low95)
```


## Updating all tests

**First calculate the updated standard deviation**

```{r}
## Calculate the updated SD based on sample size
df2 <- df %>%
  mutate(bayes_sd = sqrt(1 / ((test - 1 + 1) * 1 / prior_repeated_measure_sd^2 + 1 / prior_sd^2))) 

df2
```


**Write a for() loop to iteratively update the mean and standard deviation following each test**

```{r}
# Create a vector to store results
N <- length(df2$ln_value) + 1
bayes_mu <- c(prior_mu, rep(NA, N - 1))


## For loop
for(i in 2:N){
  
  ## Set up vectors for the variance, denominator, and newly observed values
  prior_var <- c(prior_sd^2, df2$bayes_sd^2)
  denominator <- prior_repeated_measure_sd^2 + prior_var
  vals <- df2$ln_value
  
  ## calculate bayesian updated mu
  bayes_mu[i] <- (prior_repeated_measure_sd^2 * bayes_mu[i-1] + prior_var[i-1] * vals[i-1]) / denominator[i-1]
    
}

df2$bayes_mean <- bayes_mu[-1]
df2

```


**Calculate the Confidence Intervals**

```{r}
### Confidence Intervals
first_prior <- data.frame(test = 0, value = NA, ln_value = NA, bayes_sd = prior_sd, bayes_mean = prior_mu)

df2 <- df2 %>%
  bind_rows(first_prior) %>%
  arrange(test)

## Exponentiate back to get the reference range
df2$low95 <- exp(df2$bayes_mean - 1.96*df2$bayes_sd)
df2$high95 <- exp(df2$bayes_mean + 1.96*df2$bayes_sd)
df2


```


**Create a plot to show how the reference range changes based on continuously updating our knowledge after each observation**

```{r}
### Plot ####
df2 %>%
  ggplot(aes(x = test, y = value)) +
  geom_point(size = 5, color = "blue") +
  geom_smooth(aes(y = low95),
              se = FALSE,
              color = "black",
              size = 1.2) +
  geom_smooth(aes(y = high95),
              se = FALSE,
              color = "black",
              size = 1.2) +
  ylim(0, 950)

```

